\documentclass[11pt,a4paper]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[french]{babel}
\usepackage{geometry}
\geometry{top=2.2cm,bottom=2.2cm,left=1.8cm,right=1.8cm}
\usepackage{lmodern}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{enumitem}
\setlength{\parskip}{4pt}
\usepackage{titlesec}
\usepackage{fancyhdr}
\usepackage{hyperref}
\hypersetup{colorlinks=false,pdfborder={0 0 0}}
\usepackage{array}
\usepackage{microtype}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,shapes.geometric}
\usepackage{pifont}
\usepackage{makecell}
\usepackage{caption}
\usepackage{float}
\setlength{\emergencystretch}{2.5em}
% ---------- typography ----------
\titleformat{\section}
  {\normalsize\bfseries\uppercase}
  {\thesection.}{0.5em}{}[\vspace{-2pt}\rule{\linewidth}{0.4pt}\vspace{2pt}]
\titleformat{\subsection}
  {\normalsize\bfseries}{\thesubsection.}{0.5em}{}
\titlespacing{\section}{0pt}{10pt}{4pt}
\titlespacing{\subsection}{0pt}{7pt}{2pt}

% ---------- header ----------
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0.4pt}
\fancyhead[L]{\small\textit{Research Statement --- CIFRE PhD}}
\fancyhead[R]{\small\textit{Yassine Mannai}}
\fancyfoot[C]{\small\thepage}

% ---------- theorem-like envs ----------
\newtheorem{rqthm}{Research Question}

\begin{document}

% ================================================================
%  TITLE BLOCK
% ================================================================
\begin{center}
{\Large\bfseries Research Statement}\\[4pt]
{\large\bfseries CIFRE PhD --- CREST - IMT, Generali France}\\[8pt]
{\normalsize\itshape
Regime-Conditional Generative Models, Volterra Capital Proxies\\
and Mean Field Game Lapse Dynamics for Insurance Asset Allocation}
\end{center}

\medskip
\noindent
\begin{tabular}{@{}l>{\raggedright\arraybackslash}p{11.8cm}@{}}
\textbf{Candidate:}            & Yassine Mannai \\
\textbf{Contact:}              & \texttt{yassine.mannai@dauphine.eu}
                                 $\;\cdot\;$ +33\,(0)6\,23\,64\,31\,80 \\
\textbf{Programme:}            & MSc in Quantitative Finance (272),
                                 Paris Dauphine-PSL \\
\textbf{Academic supervisors:} & Caroline Hillairet (CREST, ENSAE Paris) and \\
                               & Anthony R\'{e}veillac (INSA Toulouse, IMT) \\
\textbf{Industry supervisor:}  & Mamy Ramamonjisoa ---
                                 SAA team, Generali France \\
\textbf{Investment Division:}  & Cedrik d'Aviau de Ternay --- Chief Investment Officer, Generali France \\
\textbf{Start date:}           & January 2027
                                 $\;\cdot\;$ Duration: 36 months
\end{tabular}

\noindent\rule{\linewidth}{0.8pt}

\medskip
\noindent\textit{\textbf{Thesis in one sentence.}}
\begin{quote}\itshape
This thesis introduces the first \emph{regime-conditional generative framework
with certified $W_2$ guarantees under online regime inference}, coupled to a
\emph{differentiable functional capital proxy}, to enable a fully dynamic,
capital-aware Total Portfolio Approach for a life insurer under Solvency~II.
The three axes form a single closed control loop: a certified scenario
generator (Axis~1) feeds a fast differentiable SCR proxy (Axis~2), whose
gradient constraint drives the MFG-based lapse modelling and capital
allocation programme (Axis~3).
\end{quote}

% ================================================================
\section{Industrial Context and Motivation}
% ================================================================

\subsection{The 2022 dependence crisis and the limits of static allocation}

The year 2022 was a structural stress test for insurance asset managers: unlike
earlier single-asset shocks, equities and bonds fell simultaneously and
diversification broke down. For a life insurer such as Generali France, this
exposed a structural flaw in Strategic Asset Allocation (SAA), which assumes
stable, linear dependence between economic factors. The equity/bond correlation
switched from $-0.3$ (diversifying) to $+0.6$ (concentration risk) within
weeks --- a regime reversal that fixed-correlation models cannot anticipate.

This motivates the central problem of this thesis:
\textit{how to characterise and optimally control the solvency position of an insurer
when the joint dynamics of economic factors are governed by a latent Markov chain
inducing non-stationary, non-linear dependence structures across regimes?}

\subsection{Three compounded bottlenecks in the current SAA process}

The SAA process at Generali France runs in three sequential steps, each with
structural limitations.
\textbf{(1)~Economic Scenario Generator (ESG)} --- a proprietary ESG produces
thousands of trajectories (rates, spreads, equity, real estate, inflation)
calibrated with \textbf{fixed correlations averaged across regimes}, feeding all
downstream actuarial calculations.
\textbf{(2)~SCR via nested LSMC} --- the Solvency Capital Requirement, the
$99.5\%$ one-year VaR of net asset value change, requires the Best Estimate of
Liabilities (BEL) under $10{,}000+$ scenarios; Least-Squares Monte Carlo
(LSMC, Krah et al., 2018) replaces the inner simulation by an offline
polynomial proxy.
\textbf{(3)~Annual Markowitz optimisation} --- the SCR is a fixed constraint in
a mean-variance optimisation revised \textbf{once per year}, with no automatic
identification of the dominant portfolio objective (duration, inflation, carry,
capital efficiency), even though the relevant factor varies across portfolios
and regimes.

\medskip
\noindent\textbf{Four structural limitations.}

\begin{enumerate}[leftmargin=1.8em,itemsep=2pt]
\item \textbf{Regime-blind dependence.}
Fixed correlations cannot anticipate regime reversals: a Markowitz optimum for
a low-inflation regime becomes high-risk under stagflation.

\item \textbf{LSMC proxy limitations.}
The polynomial proxy is accurate only within its regime-agnostic calibration
region, degrading precisely in the stress zones most relevant for capital
decisions, and is not expressed in common economic factors.

\item \textbf{Static, siloed decision-making.}
Allocation and hedging are decided separately --- suboptimal for an insurer with
convex critical losses --- and the objective is imposed uniformly rather than
inferred from each portfolio's dominant risk factor.

\item \textbf{Private asset valuation bias.}
Private equity, private debt and real estate NAVs are smoothed quarterly: in
2008 private equity NAVs stayed stable for $6$--$12$ months before correcting
sharply, understating stress-regime dependence and introducing a lag bias in
the SCR.
\end{enumerate}

\subsection{The Total Portfolio Approach: target framework}

The \textbf{Total Portfolio Approach (TPA)} evaluates each decision through its
marginal contribution to total balance-sheet risk via common factors
(Duration, Inflation, Growth, Liquidity, Carry). Implementations at sovereign
wealth funds (Future Fund, CPP Investments) report about $+1\%$ excess return
per year over classical SAA (Thinking Ahead Institute, 2025; CAIA, 2024). For
an insurer under Solvency~II, the factor space must also include liability-side
factors (surrender, longevity, yield-curve/BEL interactions), and every
decision must internalise a capital cost. Existing actuarial tools address none
of this jointly; this thesis fills the gap along three coupled axes:
regime-conditional factor simulation (Axis~1), fast capital proxy (Axis~2),
and MFG-based lapse modelling and capital allocation (Axis~3).

% ================================================================
\section{Research Overview}
% ================================================================

\begin{figure}[H]
\centering
\begin{tikzpicture}[
  node distance=6mm,
  box/.style={draw,rounded corners,align=center,inner sep=4pt,
              minimum height=12mm,font=\small},
  ax/.style={box,minimum width=37mm},
  arr/.style={-{Latex[length=2mm]},thick},
  lbl/.style={font=\scriptsize\itshape,align=center}
]
\node[ax] (a1) {\textbf{Axis~1}\\Regime-conditional\\generative model\\\scriptsize(HMM-SMC + GAN/VAE/SGM)};
\node[ax,right=17mm of a1] (a2) {\textbf{Axis~2}\\Differentiable\\SCR proxy\\\scriptsize(Volterra functional)};
\node[ax,right=17mm of a2] (a3) {\textbf{Axis~3}\\Network lapse risk\\\& capital allocation\\\scriptsize(MFG on network)};
\draw[arr] (a1) -- node[lbl,above]{certified\\scenarios ($W_2$)} (a2);
\draw[arr] (a2) -- node[lbl,above]{capital\\constraint ($\nabla$SCR)} (a3);
\draw[arr] (a3.south) -- ++(0,-7mm) -| node[lbl,below,pos=0.25]{realised regime / rebalancing feedback} (a1.south);
\end{tikzpicture}
\caption{The three axes form a single closed capital-aware control loop:
certified regime-conditional scenarios feed a fast differentiable capital
proxy, whose gradient drives the capital allocation programme;
realised regimes feed back into the generator.}
\label{fig:pipeline}
\vspace{4pt}

\renewcommand{\arraystretch}{1.15}
\begin{tabularx}{\linewidth}{@{}Xcccc@{}}
\toprule
\textbf{Approach} & \makecell{Regime-\\aware} & \makecell{Differen-\\tiable} &
\makecell{$W_2$/$L^2$\\certified} & \makecell{Insurance-\\calibrated} \\
\midrule
Static SAA / Markowitz                       & \ding{55} & \ding{55} & \ding{55} & \ding{51} \\
Copulas (Sklar, Fermanian)                   & \ding{55} & \ding{55} & \ding{51} & \ding{55} \\
LSMC proxy (Krah et al.)                     & \ding{55} & \ding{55} & \ding{51} & \ding{51} \\
GAN/VAE time-series (Wiese, Yoon)            & \ding{55} & \ding{51} & \ding{55} & \ding{55} \\
Score-based models (Ocello et al.)           & \ding{55} & \ding{51} & \ding{51} & \ding{55} \\
Classical lapse models (actuarial)          & \ding{55} & \ding{55} & \ding{55} & \ding{51} \\
\midrule
\textbf{This thesis (Axes 1--3)}            & \ding{51} & \ding{51} & \ding{51} & \ding{51} \\
\bottomrule
\end{tabularx}
\captionof{table}{Positioning of the research programme against the state of the art.\\
\ding{51}~= requirement met, \ding{55}~= not addressed.}
\label{tab:positioning}
\end{figure}

% ================================================================
\section{Research Programme}
% ================================================================

% ----------------------------------------------------------------
\subsection{Axis~1 --- Regime-Conditional Factor Dependence}
% ----------------------------------------------------------------

\noindent\textbf{Research question.}
\begin{rqthm}
How to generate the joint distribution of TPA factors conditionally on a
sequentially estimated macroeconomic regime, while providing: (i)~formal
goodness-of-fit guarantees on the learned dependence structure; and
(ii)~non-asymptotic $W_2$-convergence bounds under particle-filter
inference error on the regime state?
\end{rqthm}

\noindent\textbf{State of the art.}

\textit{Regime identification.}
Hidden Markov Models (Hamilton, 1989) and Sequential Monte Carlo (SMC) particle
filters (Chopin \& Papaspiliopoulos, 2020) provide tractable real-time estimation
of latent macroeconomic regimes $z_t \in \{1,\ldots,K\}$. Their combination ---
an online HMM-SMC filter --- yields a posterior $p(z_t|\mathcal{F}_t)$ updated at
each observation.

\textit{Generative Adversarial Networks.}
Applied to financial series by Wiese et al.\ (2020, Quant-GAN) and Yoon et al.\
(2019, TimeGAN), conditional GANs generate
$\mathbf{F}_{t+1} | z_t = k \sim G_\theta(\varepsilon,k)$ with full flexibility
(no parametric dependence assumption), but suffer from \textit{mode collapse}
and the absence of distributional guarantees, limiting regulatory auditability.

\textit{Variational Autoencoders.}
CVAEs encode each trajectory into a regime-conditional posterior and maximise
the ELBO; they offer stable training and an auditable latent structure, but
produce overly smooth samples due to the Gaussian posterior approximation.

\textit{Score-Based Generative Models (SGMs).}
Strasman, Ocello et al.\ (TMLR, 2024) establish $W_2$ convergence bounds:
\begin{equation}
  \mathrm{KL}(\hat{p}\,\|\,p^*)
  \;\le\; \int_0^T g(t)^2\,\varepsilon_{\mathrm{score}}(t)\,dt
  + \varepsilon_{\mathrm{init}}.
\label{eq:kl}
\end{equation}
Gentiloni-Silveri \& Ocello (ICML, 2025) extend these guarantees to
non-log-concave distributions (Gaussian mixtures), directly applicable to
multi-regime market models. De Marco, Pham \& Zanni (2025) further introduce
jump-diffusion Schr\"{o}dinger bridges to capture discontinuous regime
transitions.

\textit{Copulas: a validation framework, not a generative model.}
Sklar's theorem separates margins from dependence, and Fermanian (2013, 2022)
provides goodness-of-fit tests for conditional copulas. Classical parametric
families (Gaussian, Student-$t$, Archimedean), however, impose rigid tails,
scale poorly in high dimension and cannot capture cross-regime dynamics.
Copulas therefore serve here exclusively as a \emph{validation tool}: the
conditional tail-dependence coefficient $\lambda_k$ is estimated and tested
via Fermanian's specification tests to validate generative outputs, not to
generate them.

\noindent\textbf{Identified gaps.}
Neither conditional GANs, VAEs nor SGMs have been applied to the joint
distribution of TPA factors with real-time regime identification.
For the GAN/VAE: the absence of formal regime-conditional tail-dependence
validation. For the SGM: the absence of $W_2$ bounds for the conditional case
with sequentially estimated regime states. These constitute the theoretical
contribution of Axis~1.

\noindent\textbf{Proposed approach.}

\textit{Stage 1 (Year 1) --- cGAN and cVAE.}
Let $z_t$ be identified by an online HMM-SMC filter. A conditional GAN
$G_\theta(\varepsilon,k)$ and a CVAE are trained in parallel, both incorporating
the private asset vintage lag factor
$\tilde{F}^{\mathrm{PA}}_t = \mathrm{NAV}^{\mathrm{PA}}_{t-1}$
to capture the quarterly smoothing bias. Model comparison is conducted on:
Wasserstein-2 distance $W_2$, Maximum Mean Discrepancy (MMD), per-regime
tail calibration via Fermanian's $\lambda_k$ tests, autocorrelation structure,
and regulatory auditability.

\textit{Stage 2 (Year 2) --- Conditional SGM with $W_2$ guarantees.}
A conditional OU forward process for regime $k$:
\begin{equation}
  d\mathbf{x}^\tau = -\mathbf{x}^\tau\,d\tau
  + \sqrt{2}\,g_k(\tau)\,dW_\tau, \quad \tau \in [0,T],
\label{eq:sgm}
\end{equation}
with regime-specific noise schedule $g_k(\tau)$. The theoretical contribution
is the extension of the $W_2$ bounds of Gentiloni-Silveri \& Ocello (ICML 2025)
to the conditional case with online SMC regime estimation --- quantifying how
inference error in $p(z_t|\mathcal{F}_t)$ propagates into the generative bound.
A scenario selection layer then extracts a central regime-conditional scenario
for downstream reporting and optimisation diagnostics.

% ----------------------------------------------------------------
\subsection{Axis~2 --- Supervised Capital Proxy}
% ----------------------------------------------------------------

\noindent\textbf{Research question.}
\begin{rqthm}
How to construct a proxy for the insurance SCR --- expressed as a function of
TPA factors, the current regime, and liability states including private asset
NAV lags formalised as a Volterra functional --- that is accurate in
decision-relevant zones with controlled $L^2$ error bounds per regime and
sub-10ms inference time, and exposes a differentiable interface for its direct
embedding as an explicit capital constraint in a downstream stochastic
optimisation programme?
\end{rqthm}

\noindent\textbf{State of the art.}

\textit{Classical proxy models.}
The LSMC framework (Krah et al., 2018) approximates the BEL by polynomial
regression on a design-of-experiment. Its two critical limitations are
regime-blindness and degraded accuracy in stress zones. Neural network proxies
(Huber \& Schmocker, 2022) achieve $1$--$3\%$ accuracy at $<10$ms inference.
Bourgey (2020) provides theoretical foundations via multi-level Monte Carlo
(MLMC) estimators for nested expectations and polynomial chaos expansions with
$L^2$ error estimates, motivating the adaptive-weighting architecture of
Axis~2.

\textit{Recent generative and flow-based surrogates.}
Three recent architectures merit systematic comparison for their
uncertainty-quantification and stability properties:
\textbf{Neural ODEs} (Chen et al., 2018), which model the proxy surface as a
continuous-depth dynamical system; \textbf{Conditional Flow Matching} (Lipman
et al., 2022), which learns straight-line ODE paths offering stable training
and a full conditional density
$p(\widehat{\mathrm{SCR}} | \mathbf{F}, z, \mathbf{P})$ for uncertainty-aware
allocation; and \textbf{score-based diffusion surrogates}. \textbf{Tabular
diffusion models} (TabDDPM, Kotelnikov et al., 2023) additionally serve as
design-of-experiment augmentation for under-represented regimes.

\textit{Volterra processes and path-dependent risk measures.}
The private asset NAV lag is not a Markovian state variable. The observed
smoothed value
$\widetilde{\mathrm{NAV}}_t = \int_0^t K(t,s)\,\mathrm{NAV}^*_s\,ds$
is a Volterra functional of the true (unobservable) NAV process, where $K(t,s)$
is a smoothing kernel. The theory of affine Volterra processes (Abi Jaber,
Larsson \& Pulido, 2019; El Euch \& Rosenbaum, 2019) provides a tractable
framework for such path-dependent dynamics. This motivates extending the
capital risk measure to the functional setting:
\begin{equation}
  \rho^V(L_T)
  = \phi\!\left(\int_0^T K(T,s)\,\mathrm{CVaR}_\alpha(L_s)\,ds\right),
\label{eq:volterra}
\end{equation}
where $\phi$ is a monetary risk measure. Equation~\eqref{eq:volterra}
formalises the temporally diffused risk crystallisation of private assets and
provides a theoretical grounding for the vintage lag correction embedded in the
proxy state vector.

\textit{Model comparison and interoperability metrics.}
Proxy architectures will be evaluated on two sets of criteria.
\textit{Comparison metrics}: (i)~accuracy --- RMSE, relative error at the SCR
regulatory threshold, $L^\infty$ norm in stress zones; (ii)~calibration ---
prediction-interval coverage, reliability diagrams; (iii)~speed --- inference
latency (target $<5$ms) and batch throughput; (iv)~interpretability ---
SHAP values, integrated gradients, first-order Sobol sensitivity indices
$S_i = \mathrm{Var}[\mathbb{E}[\widehat{\mathrm{SCR}}|\mathbf{F}_i]]/
\mathrm{Var}[\widehat{\mathrm{SCR}}]$.
\textit{Interoperability metrics}: (i)~differentiability ---
$\nabla_\mathbf{F}\widehat{\mathrm{SCR}}$ available via automatic
differentiation for gradient-based embedding in the Axis~3 optimiser;
(ii)~API compatibility with GPS and PyTorch/JAX optimisation frameworks;
(iii)~monotonicity and convexity preservation as required by Solvency~II
constraints; (iv)~memory footprint for production deployment.

\noindent\textbf{Identified gaps.}
No existing proxy is jointly conditioned on the macroeconomic regime, TPA
factors and private asset NAV lag. None accounts for the path-dependent
Volterra structure of NAV smoothing, and none provides the differentiable
interface required for its direct embedding as an explicit capital constraint
in a dynamic optimisation programme.

\noindent\textbf{Proposed approach.}
We learn the supervised proxy:
\begin{equation}
  \widehat{\mathrm{SCR}}(t)
  = f_\theta\!\bigl(\mathbf{F}_t,\,z_t,\,\mathbf{P}_t\bigr),
\label{eq:proxy}
\end{equation}
where $\mathbf{P}_t$ includes effective duration, average guaranteed rate,
observed lapse ratio and the vintage lag $\tilde{F}^{\mathrm{PA}}_{t-1}$.
The network $f_\theta$ is trained with \textbf{adaptive sample weighting}
concentrating mass near the regulatory threshold and in stress regimes.
The generalisation bound:
\begin{equation}
  \bigl|\mathrm{SCR}(t) - f_\theta(\mathbf{F}_t,z_t,\mathbf{P}_t)\bigr|
  \;\le\; \varepsilon_k(n_k,\delta),
\label{eq:bound}
\end{equation}
decreasing with $n_k$ (training points in regime $k$) and with confidence
$1-\delta$. The four architectures (MLP+attention, Neural ODE, Flow Matching,
score-based) are compared on the full metric set above, and the selected proxy
is exported as a differentiable module ($\nabla_\mathbf{F}\widehat{\mathrm{SCR}}$
available) for direct use as an explicit constraint in the Axis~3 programme.

% ----------------------------------------------------------------
\subsection{Axis~3 --- Mean Field Game Network Model for Lapse Risk
  and Capital Allocation}
% ----------------------------------------------------------------

\noindent\textbf{Research question.}
\begin{rqthm}
How to model policyholder surrender dynamics as a mean field game on a
heterogeneous policy network conditioned on the macroeconomic regime of
Axis~1, derive the regime-conditional aggregate lapse process and its
path-dependent impact on the SCR proxy of Axis~2, and solve the resulting
capital allocation problem under a Solvency~II shortfall constraint?
\end{rqthm}

\noindent\textbf{State of the art.}

\textit{Classical lapse models.}
Current actuarial practice models surrender rates as deterministic functions
of the interest-rate spread, calibrated on historical cohorts (Kochanski,
2010 style). Dynamic models incorporate rate-sensitivity but assume
independence across policyholders --- ignoring social contagion. In 2022,
French life insurers observed correlated surrender waves far in excess of
individual rational behaviour, motivating a network-contagion approach.

\textit{Mean field games.}
Lasry \& Lions (2007) and Carmona \& Delarue (2018) establish MFG theory
for large populations of interacting agents: each agent solves an HJB
equation whose coefficients depend on the aggregate distribution $m_t$
(mean field), yielding a Nash equilibrium characterised by a coupled
forward-backward PDE system. MFGs have been applied to systemic risk in
banking (Carmona et al., 2015) but not to insurance lapse contagion.

\textit{Financial network contagion.}
Bardoscia et al.\ (2021) establish that network topology --- degree
distribution, community structure, contagion thresholds --- determines
systemic amplification. Applied to a policyholder network (nodes = contracts,
edges = social or geographic proximity), the same framework governs surrender
cascades.

\textit{Stochastic control and capital allocation.}
R\'{e}veillac (2013) and Pham (2009) provide the BSDE and HJB framework
for optimal allocation under dynamic risk constraints. Rosazza Gianin et al.\
(2024) supply axiomatic capital allocation rules under shortfall risk measures,
directly applicable once the MFG lapse process is embedded as a liability-side
state variable.

\noindent\textbf{Identified gaps.}
No existing model treats policyholder lapse as a MFG equilibrium on a
network conditioned on the macroeconomic regime. No framework connects the
aggregate lapse process to a differentiable SCR proxy or to a Solvency~II
capital allocation programme.

\noindent\textbf{Proposed approach.}
Each policyholder $i$ is a MFG agent with in-force indicator $x_i(t)$ and
value function $V^i$ satisfying
\begin{equation}
  -\partial_t V^i
  = \min_{\alpha_i}\Bigl[\ell(x_i,\alpha_i,m_t,z_t)
    + \mathcal{L}^{\alpha_i}V^i\Bigr],
  \quad m_t = \tfrac{1}{N}\textstyle\sum_j\delta_{x_j(t)},
\label{eq:mfg}
\end{equation}
where $z_t$ is the regime from Axis~1 and local interaction terms encode
the network topology. The MFG equilibrium yields a regime-conditional
aggregate lapse rate $\lambda^*(t,z_t,m_t)$ that enters the liability
state vector $\mathbf{P}_t$ of the Axis~2 proxy. The capital allocation
problem then reads
\begin{equation}
  \min_{\mathbf{w}}\;\mathrm{CVaR}_{0.99}(\Delta\mathrm{NAV}_T)
  \quad\text{s.t.}\quad
  \widehat{\mathrm{SCR}}\!\bigl(\mathbf{F}_t,z_t,\mathbf{P}_t(\lambda^*)\bigr)
  \le \bar{C},
\label{eq:alloc}
\end{equation}
where the differentiable proxy of Axis~2 enables gradient-based resolution
of~\eqref{eq:alloc}. The MFG system is solved via deep Galerkin methods
(physics-informed NNs) and validated on the 2022 French surrender-wave
episode.


% ================================================================
\section{Timeline (36 Months)}
% ================================================================

\begin{itemize}[leftmargin=1.6em,itemsep=2pt]
  \item \textbf{Year~1 --- Axis~1.}
        Bloomberg factor database with NAV lags; online HMM-SMC calibration;
        regime-conditional GAN/CVAE validated on 2008/2020/2022 crises.
        \ Publication in mathematical finance / statistics.
  \item \textbf{Year~2 --- Axis~2.}
        Actuarial design-of-experiment (Prophet/MoSes) with TabDDPM
        augmentation; SCR proxy architecture comparison and Volterra $L^2$
        bounds; differentiable proxy embedded as optimisation constraint.
        \ Publication in actuarial science / insurance mathematics.
  \item \textbf{Year~3 --- Axis~3.}
        MFG-network lapse calibration on the 2022 French surrender-wave
        episode; full three-axis coupling; capital allocation out-of-sample
        backtesting; defence.
        \ Publication in quantitative finance.
\end{itemize}

% ================================================================
\section{Conclusion}
% ================================================================

This thesis delivers a tightly coupled three-module system designed to make the
\textbf{Total Portfolio Approach operational for a life insurer under
Solvency~II}, with direct integration into Generali France's proprietary
platform GPS:

\begin{itemize}[leftmargin=1.6em,itemsep=3pt]
  \item \textbf{Module 1 (Axis~1)} replaces static ESG correlations with an
        auditable regime-conditional generative model, enabling real-time
        scenario generation that reflects the actual dependence structure of
        the prevailing macroeconomic regime, including the path-dependent
        correction dynamics of private assets.

  \item \textbf{Module 2 (Axis~2)} replaces the multi-hour SCR calculation
        with a $<5$ms differentiable proxy grounded in Volterra functional
        theory, making dynamic portfolio optimisation loops computationally
        feasible and the capital constraint analytically tractable as an
        explicit gradient-available term.

  \item \textbf{Module 3 (Axis~3)} replaces deterministic lapse tables
        with a mean field game equilibrium on the policyholder network,
        delivering a regime-conditional aggregate lapse process that feeds
        directly into the SCR proxy and enables a capital allocation
        programme with explicit Solvency~II guarantees.
\end{itemize}

Beyond the industrial deliverables, the thesis makes three mathematically
self-contained contributions: the extension of $W_2$ convergence theory to
regime-conditional score-based models under sequential estimation error; the
formalisation of private asset valuation bias through the theory of Volterra
processes and path-dependent risk measures; and the characterisation of policyholder surrender equilibria as mean field
games on network-structured populations with regime-conditional interaction
kernels, and their integration into Solvency~II capital allocation programmes
via the differentiable SCR proxy of Axis~2. Together,
these contributions establish rigorous foundations for a new generation of
quantitative tools in insurance balance sheet management --- tools that are
simultaneously \emph{theoretically grounded}, \emph{computationally efficient},
\emph{interoperable} and \emph{regulatory-compliant}.

% ================================================================
\section*{References}
% ================================================================
\renewcommand{\section}[2]{}

\begin{thebibliography}{26}

\bibitem{abijabar2019}
Abi Jaber, E., Larsson, M., \& Pulido, S. (2019).
Affine Volterra processes.
\emph{The Annals of Applied Probability}, 29(6), 3155--3200.

\bibitem{bardoscia2021}
Bardoscia, M., Barucca, P., Battiston, S., Caccioli, F., Cimini, G.,
Garlaschelli, D., Saracco, F., Squartini, T., \& Caldarelli, G. (2021).
The physics of financial networks.
\emph{Nature Reviews Physics}, 3(7), 490--507.

\bibitem{bourgey2020}
Bourgey, F. (2020).
\emph{Stochastic approximations for financial risk computations}.
PhD thesis.

\bibitem{caia2024}
CAIA Association. (2024).
\emph{Innovation Unleashed: The Rise of Total Portfolio Approach}.

\bibitem{carmona2018}
Carmona, R., \& Delarue, F. (2018).
\emph{Probabilistic Theory of Mean Field Games with Applications}.
Springer.

\bibitem{chen2018}
Chen, R.T.Q., Rubanova, Y., Bettencourt, J., \& Duvenaud, D. (2018).
Neural ordinary differential equations.
\emph{Advances in Neural Information Processing Systems}, 31.

\bibitem{chopin2020}
Chopin, N., \& Papaspiliopoulos, O. (2020).
\emph{An Introduction to Sequential Monte Carlo}. Springer.

\bibitem{demarco2025}
De Marco, S., Pham, H., \& Zanni, D. (2025).
Schr\"{o}dinger bridges with jumps for time series generation.
\emph{arXiv preprint arXiv:2602.20011}.

\bibitem{eleuch2019}
El Euch, O., \& Rosenbaum, M. (2019).
The characteristic function of rough Heston models.
\emph{Mathematical Finance}, 29(1), 3--38.

\bibitem{fermanian2013}
Fermanian, J.-D. (2013).
An overview of the goodness-of-fit test problem for copulas.
In \emph{Copulae in Mathematical and Quantitative Finance},
Springer, 61--89.

\bibitem{fermanian2022}
Fermanian, J.-D. (2022).
Conditional copulas, dependence modeling and the role of conditioning.
\emph{Annals of Statistics}.

\bibitem{ocello2025icml}
Gentiloni-Silveri, M., \& Ocello, A. (2025).
Beyond log-concavity and score regularity: Improved convergence bounds
for score-based generative models in $W_2$-distance.
\emph{Forty-Second International Conference on Machine Learning (ICML)}.

\bibitem{hamilton1989}
Hamilton, J.D. (1989).
A new approach to the economic analysis of nonstationary time series.
\emph{Econometrica}, 57(2), 357--384.

\bibitem{huber2022}
Huber, M., \& Schmocker, B. (2022).
A neural network approach to efficient valuation of large portfolios
of variable annuities.
\emph{Insurance: Mathematics and Economics}, 105, 96--116.

\bibitem{kotelnikov2023}
Kotelnikov, A., Baranchuk, D., Rubachev, I., \& Babenko, A. (2023).
TabDDPM: Modelling Tabular Data with Diffusion Models.
\emph{International Conference on Machine Learning (ICML)}.

\bibitem{krah2018}
Krah, A.-S., Nikolic, Z., \& Korn, R. (2018).
A Least-Squares Monte Carlo framework in proxy modeling of life
insurance companies.
\emph{Risks}, 6(2), 62.

\bibitem{lasry2007}
Lasry, J.-M., \& Lions, P.-L. (2007).
Mean field games.
\emph{Japanese Journal of Mathematics}, 2(1), 229--260.

\bibitem{lipman2022}
Lipman, Y., Chen, R.T.Q., Ben-Hamu, H., Nickel, M., \& Le, M. (2022).
Flow Matching for Generative Modeling.
\emph{International Conference on Learning Representations (ICLR)}.

\bibitem{pham2009}
Pham, H. (2009).
\emph{Continuous-Time Stochastic Control and Optimization
with Financial Applications}. Springer.

\bibitem{reveillac2013}
R\'{e}veillac, A. (2013).
CRRA utility maximization under dynamic risk constraints.
\emph{Mathematical Finance}.

\bibitem{rosazza2024}
Rosazza Gianin, E., et al.\ (2024).
Shortfall systemic risk measures, capital allocation rules and
generalised collapse to the mean: Theory and practice.
\emph{Working paper}.

\bibitem{sklar1959}
Sklar, A. (1959).
Fonctions de repartition a $n$ dimensions et leurs marges.
\emph{Publications de l'Institut de Statistique de Paris}, 8, 229--231.

\bibitem{ocello2024tmlr}
Strasman, S., Ocello, A., Boyer, C., Le~Corff, S., \& Lemaire, V. (2024).
An analysis of the noise schedule for score-based generative models.
\emph{Transactions on Machine Learning Research (TMLR)}.

\bibitem{thinking_ahead}
Thinking Ahead Institute. (2025).
\emph{TPA Report: From Vision to Execution}. WTW / CAIA.

\bibitem{wiese2020}
Wiese, M., Knobloch, R., Korn, R., \& Kretschmer, P. (2020).
Quant GANs: Deep generation of financial time series.
\emph{Quantitative Finance}, 20(9), 1419--1440.

\bibitem{yoon2019}
Yoon, J., Jarrett, D., \& van der Schaar, M. (2019).
Time-series generative adversarial networks.
\emph{Advances in Neural Information Processing Systems}, 32.

\end{thebibliography}

\end{document}
Abi Jaber, E., Larsson, M., \& Pulido, S. (2019).
Affine Volterra processes.
\emph{The Annals of Applied Probability}, 29(6), 3155--3200.

\bibitem{bardoscia2021}
Bardoscia, M., Barucca, P., Battiston, S., Caccioli, F., Cimini, G.,
Garlaschelli, D., Saracco, F., Squartini, T., \& Caldarelli, G. (2021).
The physics of financial networks.
\emph{Nature Reviews Physics}, 3(7), 490--507.

\bibitem{bourgey2020}
Bourgey, F. (2020).
\emph{Stochastic approximations for financial risk computations}.
PhD thesis.

\bibitem{buehler2019}
Buehler, H., Gonon, L., Teichmann, J., \& Wood, B. (2019).
Deep hedging.
\emph{Quantitative Finance}, 19(8), 1271--1291.

\bibitem{caia2024}

Chen, R.T.Q., Rubanova, Y., Bettencourt, J., \& Duvenaud, D. (2018).
Neural ordinary differential equations.
\emph{Advances in Neural Information Processing Systems}, 31.

\bibitem{chopin2020}
Chopin, N., \& Papaspiliopoulos, O. (2020).
\emph{An Introduction to Sequential Monte Carlo}. Springer.

\bibitem{demarco2025}
De Marco, S., Pham, H., \& Zanni, D. (2025).
Schr\"{o}dinger bridges with jumps for time series generation.
\emph{arXiv preprint arXiv:2602.20011}.

\bibitem{eleuch2019}
El Euch, O., \& Rosenbaum, M. (2019).
The characteristic function of rough Heston models.
\emph{Mathematical Finance}, 29(1), 3--38.

\bibitem{fermanian2013}
Fermanian, J.-D. (2013).
An overview of the goodness-of-fit test problem for copulas.
In \emph{Copulae in Mathematical and Quantitative Finance},
Springer, 61--89.

\bibitem{fermanian2022}
Fermanian, J.-D. (2022).
Conditional copulas, dependence modeling and the role of conditioning.
\emph{Annals of Statistics}.

\bibitem{ocello2025icml}
Gentiloni-Silveri, M., \& Ocello, A. (2025).
Beyond log-concavity and score regularity: Improved convergence bounds
for score-based generative models in $W_2$-distance.
\emph{Forty-Second International Conference on Machine Learning (ICML)}.

\bibitem{goodfellow2014}
Goodfellow, I. et al. (2014).
Generative adversarial networks.
\emph{Advances in Neural Information Processing Systems}, 27.

\bibitem{hamilton1989}

\bibitem{reveillac2013}
R\'{e}veillac, A. (2013).
CRRA utility maximization under dynamic risk constraints.
\emph{Mathematical Finance}.

\bibitem{rosazza2024}
Rosazza Gianin, E., et al.\ (2024).
Shortfall systemic risk measures, capital allocation rules and
generalised collapse to the mean: Theory and practice.
\emph{Working paper}.

\bibitem{sklar1959}
Sklar, A. (1959).
Fonctions de repartition a $n$ dimensions et leurs marges.
\emph{Publications de l'Institut de Statistique de Paris}, 8, 229--231.

\bibitem{ocello2024tmlr}
Strasman, S., Ocello, A., Boyer, C., Le~Corff, S., \& Lemaire, V. (2024).
An analysis of the noise schedule for score-based generative models.
\emph{Transactions on Machine Learning Research (TMLR)}.

\bibitem{thinking_ahead}
Thinking Ahead Institute. (2025).
\emph{TPA Report: From Vision to Execution}. WTW / CAIA.

\bibitem{wekwete2023}
Wekwete, T. (2023).
Application of Deep Reinforcement Learning in Asset Liability Management.
\emph{Actuarial Society of South Africa Convention Paper}.

\bibitem{wiese2020}
Wiese, M., Knobloch, R., Korn, R., \& Kretschmer, P. (2020).
Quant GANs: Deep generation of financial time series.
\emph{Quantitative Finance}, 20(9), 1419--1440.

\bibitem{yoon2019}
Yoon, J., Jarrett, D., \& van der Schaar, M. (2019).
Time-series generative adversarial networks.
\emph{Advances in Neural Information Processing Systems}, 32.

\end{thebibliography}

\end{document}
